Technique for \earlywarning\ filtering for compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-28 introduction.tex
lloid_paper: polished EM counterpart section
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1175 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a neutron star (\textsc{ns}) is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  The gravitational-wave signal is in principle detectable
before merger.  Effort from both the \GW{} and astronomy communities may make
it possible to use \GW{} observations as an early warning trigger for \EM{}
followup. In the first generation of ground-based laser interferometers, the \GW{}
community initiated a project to send alerts when potential \GW{} transients
were observed in order to trigger followup observations by \EM{} telescopes.
The typical latencies were 30 minutes \citep{HugheyGWPAW2011}, which was an
important achievement, but too late to catch the potential prompt \EM{}
emission. We have the ambition of reporting GW candidates not minutes after
the merger, as has already been accomplished, but seconds before.  We discuss
one essential ingredient in this effort, a computationally inexpensive
real-time filtering algorithm for detecting inspiral signals in gravitational
wave data.  We also present the prospects for advanced \GW\ detectors and
discuss other areas of work that would be required for rapid analysis.

@@

Though compact binary coalescence (\CBC{}) events may be responsible for
several classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}.
In this scenario, prompt gamma-ray emission arises on the accretion timescale
of a central compact object formed after the merger, 0.1--1.0\,s
\citep{Janka1999}.  Interestingly, roughly a quarter to half of observed short
\GRB{}s
also exhibit extended emission $\sim$$100$\,s.  Potential explanations for the
emission are delayed fall-back accretion \citep{Metzger2010} caused by
$r$-process heating or the formation of a proto-magnetar that interacts with
ejecta \citep{Bucciantini2011}.  In the case of the proto-magnetar, it might be
formed from accretion induced collapse, or the merger of two \textsc{ns}s.  An
accompanying \GW{} observation would confirm the latter.

>>>>

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a runaway inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges.  It is
thought that if a neutron star (\textsc{ns}) is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic (\EM{}) counterpart
\citep{shibata:2007}.  The \GW{} signal is in principle detectable
before merger.  Effort from both the \GW{} and astronomy communities may make
it possible to use \GW{} observations as an early warning trigger for \EM{}
followup. In the first generation of ground-based laser interferometers, the \GW{}
community initiated a project to send alerts when potential \GW{} transients
were observed in order to trigger followup observations by \EM{} telescopes.
The typical latencies were 30 minutes \citep{HugheyGWPAW2011}, which was an
important achievement, but too late to catch any prompt \EM{}
emission. We have the ambition of reporting \GW{} candidates not minutes after
the merger, as has already been accomplished, but seconds before.  We discuss
one essential ingredient in this effort, a computationally inexpensive
low-latency filtering algorithm for detecting inspiral signals in \GW{}
data.  We also present the prospects for advanced \GW\ detectors and
discuss other areas of work that would be required for rapid analysis.

@@

Though compact binary coalescence (\CBC{}) events may be responsible for
several classes of \EM{} transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s) \citep{Lee:2005, nakar07}.
In this scenario, prompt gamma-ray emission arises on the accretion timescale
of a central compact object formed after the merger, $0.1$--$1.0$\,s
\citep{Janka1999}.  Interestingly, roughly a quarter to half of observed short
\GRB{}s also exhibit extended emission of $30$--$100$\,s in duration beginning
$\sim$$10$\,s after the \GRB{} and carrying comparable fluence to the initial
outburst.  Potential explanations for the emission are delayed fall-back
accretion caused by $r$-process heating \citep{Metzger2010} or the formation of
a proto-magnetar that interacts with
ejecta \citep{Bucciantini2011}.  In the case of the proto-magnetar, it might be
formed from accretion-induced collapse, or the merger of two \textsc{ns}s.  An
accompanying \GW{} observation would confirm the latter.
